Supplementary Appendix

### 2025-07-23 paper/preprint/appendix.tex
formatting improved and toc added (to this end, turned appendix into standalone document).

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\usepackage{siunitx}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Supplementary Appendix},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training }
\title{Supplementary Appendix}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

For all generators used in this work we use standard logit crossentropy
loss for \(\text{yloss}(\cdot)\). All generators also penalize the
distance (\(\ell_1\)-norm) of counterfactuals from their original
factual state. For \emph{Generic} and \emph{ECCo}, we have
\(\mathcal{Z}:=\mathcal{X}\) and
\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
are searched directly in the feature space. Conversely, \emph{REVISE}
traverses the latent space of a variational autoencoder (VAE) fitted to
the training data, where \(g(\cdot)\) corresponds to the decoder
(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
the distance penalty, \emph{ECCo} uses an additional penalty component
that regularizes the energy associated with the counterfactual,
\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
al. 2024}).

@@

In \textbf{?@sec-constraints} we explain that to avoid penalizing
implausibility that arises due to mutability constraints, we impose a
point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
argue in \textbf{?@sec-constraints} that this approach induces models to
be less sensitive to immutable features and demonstrate this empirically
in \textbf{?@sec-experiments}. Below we derive the analytical results in
Prp.\textasciitilde{}\ref{prp-mtblty}.

@@

Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
target class. Then the partial derivative of the contrastive divergence
penalty \textbf{?@eq-div} with respect to coefficient \(\theta_{y^+,d}\)
is equal to

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]

@@

As discussed \textbf{?@sec-experiments}, CT is sensitive to certain
hyperparameter choices. We study the effect of many hyperparameters
extensively in Section~\ref{sec-app-grid}. For the main results, we tune
a small set of key hyperparameters (Section~\ref{sec-app-tune}). The
final choices for the main results are presented for each data set in
Table~\ref{tbl-final-params} along with training, test and batch sizes.

@@

\caption{\label{tbl-final-params}Final hyperparameters used for the main
results presented in \textbf{?@sec-experiments}. Any hyperparameter not
shown here is set to its default value (Note~\ref{nte-train-default}).}

@@

\caption{\label{tbl-ci}Confidence intervals}

@@

\begin{tabular}{lccccc}
  \toprule
  \textbf{Variable} & \textbf{Data} & \textbf{Full} & \textbf{Vanilla} & \textbf{LB} & \textbf{UB} \\\midrule
  Cost $(-\%)$ & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
  $ \text{IP}^* $ $(-\%)$ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
  $ \text{IP} $ $(-\%)$ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
  Cost $(-\%)$ & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\\midrule
  $ \text{IP}^* $ $(-\%)$ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
  $ \text{IP} $ $(-\%)$ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
  Cost $(-\%)$ & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
  $ \text{IP}^* $ $(-\%)$ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
  $ \text{IP} $ $(-\%)$ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
  Cost $(-\%)$ & OL & 1.62 & 2.63 & -1.15 & -0.81 \\\midrule
  $ \text{IP}^* $ $(-\%)$ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
  $ \text{IP} $ $(-\%)$ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\
  Cost $(-\%)$ & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
  $ \text{IP}^* $ $(-\%)$ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
  $ \text{IP} $ $(-\%)$ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
  Cost $(-\%)$ & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
  $ \text{IP}^* $ $(-\%)$ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
  $ \text{IP} $ $(-\%)$ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
  Cost $(-\%)$ & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
  $ \text{IP}^* $ $(-\%)$ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
  $ \text{IP} $ $(-\%)$ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
  Cost $(-\%)$ & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
  $ \text{IP}^* $ $(-\%)$ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
  $ \text{IP} $ $(-\%)$ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
  Cost $(-\%)$ & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
  $ \text{IP}^* $ $(-\%)$ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
  $ \text{IP} $ $(-\%)$ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
\end{tabular}

@@

\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
counterfactual training. Factual images are shown on the diagonal, with
the corresponding counterfactual for each target class (columns) in that
same row. The underlying generator, \emph{ECCo}, aims to generate
counterfactuals that are faithful to the model
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Generator Parameters} (Section~\ref{sec-app-grid-gen}):
  Investigates the effect of changing hyperparameters that affect the
  counterfactual outcomes during the training phase.
\item
  \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
  Investigates the effect of changing the penalty strengths in out
  proposed objective (\textbf{?@eq-obj}).
\item
  \textbf{Other Parameters} (Section~\ref{sec-app-grid-train}):
  Investigates the effect of changing other training parameters,
  including the total number of generated counterfactuals in each epoch.
\end{enumerate}

@@

To measure predictive performance, we compute the accuracy and F1-score
for all models on test data (Table~\ref{tbl-acc-gen},
Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
explanatory performance, we report here our findings for the
(im)plausibility and cost of counterfactuals at test time. Since the
computation of our proposed divergence metric (\textbf{?@eq-impl-div})
is memory-intensive, we rely on the distance-based metric for the grid
searches. For the counterfactual evaluation, we draw factual samples
from the training data for the grid searches to avoid data leakage with
respect to our final results reported in the body of the paper.
Specifically, we want to avoid choosing our default hyperparameters
based on results on the test data. Since we are optimizing for
explainability, not predictive performance, we still present test
accuracy and F1-scores.

@@

We do observe strong sensitivity to certain hyperparameters, with clear
an manageable patterns. Concerning generator parameters, we firstly find
that using \emph{REVISE} to generate counterfactuals during training
typically yields the worst outcomes out of all generators, often leading
to a substantial decrease in plausibility. This finding can be
attributed to the fact that \emph{REVISE} effectively assigns the task
of learning plausible explanations from the model itself to a surrogate
VAE. In other words, counterfactuals generated by \emph{REVISE} are less
faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
would expect them to be a less effective and, in fact, potentially
detrimental role in our training regime. Secondly, we observe that
allowing for a higher number of maximum steps \(T\) for the
counterfactual search generally yields better outcomes. This is
intuitive, because it allows more counterfactuals to reach maturity in
any given iteration. Looking in particular at the results for
\emph{Linearly Separable}, it seems that higher values for \(T\) in
combination with higher decision thresholds (\(\tau\)) yields the best
results when using \emph{ECCo}. But depending on the degree of class
separability of the underlying data, a high decision-threshold can also
affect results adversely, as evident from the results for the
\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
here we find that CT generally fails to achieve its objective because
only a tiny proportion of counterfactuals ever reaches maturity.

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]

@@

>>>>

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\usepackage{siunitx}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Supplementary Appendix},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training }
\title{Supplementary Appendix}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Counterfactual Training: Teaching Models Plausible and
Actionable Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{}
\date{}
\begin{document}
\maketitle
\begin{abstract}
This is the supplementary appendix to our paper titled
\emph{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}. It provides helpful details on mathematical notations and
formulas, our proposed training regime, extended empirical findings,
hyperparameter tuning and grid searches as well as software and
computations.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

\pagebreak
\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\newpage{}

@@

For all generators used in this work we use standard logit crossentropy
loss for \(\text{yloss}(\cdot)\). All generators also penalize the
distance (\(\ell_1\)-norm) of counterfactuals from their original
factual state. For \emph{Generic} and \emph{ECCCo}, we have
\(\mathcal{Z}:=\mathcal{X}\) and
\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
are searched directly in the feature space. Conversely, \emph{REVISE}
traverses the latent space of a variational autoencoder (VAE) fitted to
the training data, where \(g(\cdot)\) corresponds to the decoder
(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
the distance penalty, \emph{ECCCo} uses an additional penalty component
that regularizes the energy associated with the counterfactual,
\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
al. 2024}).

@@

In the main paper, we explain that to avoid penalizing implausibility
that arises due to mutability constraints, we impose a point mass prior
on \(p(\mathbf{x})\) for the corresponding feature. We argue that this
approach induces models to be relatively less sensitive to immutable
features, propose a theoretical result supporting this and provide
empirical evidence that strengthens our argument (both in the main paper
and additional findings in this appendix). Below we derive the
analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.

@@

Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
target class. Then the partial derivative of the contrastive divergence
penalty with respect to coefficient \(\theta_{y^+,d}\) is equal to

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]

@@

As discussed the main paper, CT is sensitive to certain hyperparameter
choices. We study the effect of many hyperparameters extensively in
Section~\ref{sec-app-grid}. For the main results, we tune a small set of
key hyperparameters (Section~\ref{sec-app-tune}). The final choices for
the main results are presented for each data set in
Table~\ref{tbl-final-params} along with training, test and batch sizes.

@@

\caption{\label{tbl-final-params}Final hyperparameters used for the main
results presented in the main paper. Any hyperparameter not shown here
is set to its default value (Note~\ref{nte-train-default}).}

@@

\caption{\label{tbl-ci}Mean outcomes for \textbf{CT} and \textbf{BL}
along with bootstrapped confidence intervals (99\%) for difference in
mean outcomes grouped by dataset and evaluation metric. Column
\textbf{LB} and \textbf{UB} show the lower and upper bound of the
intervals, respectively, and computed using the percentile method. The
underlying counterfactual evaluations are the same as the ones used to
produce the main table in the paper.}

@@

\begin{tabular}{lccccc}
  \toprule
  \textbf{Variable} & \textbf{Data} & \textbf{CT} & \textbf{BL} & \textbf{LB} & \textbf{UB} \\\midrule
  Cost & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
  Cost & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
  Cost & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\
  Cost & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
  Cost & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
  Cost & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
  Cost & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
  Cost & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
  Cost & OL & 1.62 & 2.63 & -1.15 & -0.81 \\
  $ \text{IP}^* $ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
  $ \text{IP}^* $ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
  $ \text{IP}^* $ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
  $ \text{IP}^* $ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
  $ \text{IP}^* $ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
  $ \text{IP}^* $ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
  $ \text{IP}^* $ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
  $ \text{IP}^* $ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
  $ \text{IP}^* $ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
  $ \text{IP} $ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
  $ \text{IP} $ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
  $ \text{IP} $ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
  $ \text{IP} $ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
  $ \text{IP} $ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
  $ \text{IP} $ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
  $ \text{IP} $ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
  $ \text{IP} $ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
  $ \text{IP} $ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\\bottomrule
\end{tabular}

@@

\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
counterfactual training. Factual images are shown on the diagonal, with
the corresponding counterfactual for each target class (columns) in that
same row. The underlying generator, \emph{ECCCo}, aims to generate
counterfactuals that are faithful to the model
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}

@@

\begin{minipage}{0.50\linewidth}

\subsubsection{Integrated Gradients}\label{integrated-gradients}

\begin{table}[H]

\caption{\label{tbl-ig}Integrated gradients.}

\begin{tabular}{
l
S[table-format=2.2(3.2)]
S[table-format=2.2(2.2)]
}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 0.03 \pm 0.0 & 10.24 \pm 2.4 \\
  Circ & 3.2 \pm 0.67 & 149.76 \pm 842.75 \\
  Moon & 60.84 \pm 128.51 & 0.55 \pm 0.06 \\
  OL & 0.78 \pm 0.12 & 4.81 \pm 1.08 \\\midrule
  Adult & 0.43 \pm 0.01 & 1.0 \pm 0.0 \\
  CH & 0.08 \pm 0.01 & 0.23 \pm 0.01 \\
  Cred & 0.0 \pm 0.0 & 0.43 \pm 0.01 \\
  GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\
\end{tabular}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}

\caption{\label{fig-mnist-ig}Class-conditional integrated gradients.}

\subsubsection{Costs and Validity}\label{costs-and-validity}

\caption{\label{tbl-panel}Costs and validity.}

\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-costs}Costs}

\begin{tabular}{
l
S[table-format=2.2(1.2)]
}
  \toprule
  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
  LS & -26.82\pm0.86 $^{*}$ \\
  Circ & 40.97\pm0.82 $^{*}$ \\
  Moon & 33.83\pm0.98 $^{*}$ \\
  OL & 10.35\pm1.28 $^{*}$ \\\midrule
  Adult & 1.16\pm3.53 $^{}$ \\
  CH & -34.89\pm2.31 $^{*}$ \\
  Cred & 28.24\pm1.08 $^{*}$ \\
  GMSC & 3.54\pm5.78 $^{}$ \\
  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
  Avg. & 2.75 \\\bottomrule
\end{tabular}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-val}Validity}

\begin{tabular}{lcc}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 1.0 & 1.0 \\
  Circ & 0.97 & 0.52 \\
  Moon & 1.0 & 1.0 \\
  OL & 0.87 & 0.98 \\\midrule
  Adult & 0.61 & 0.99 \\
  CH & 0.96 & 1.0 \\
  Cred & 0.7 & 1.0 \\
  GMSC & 0.63 & 1.0 \\
  MNIST & 1.0 & 1.0 \\
\end{tabular}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-val-mtbl}Validity}

\begin{tabular}{lcc}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 1.0 & 1.0 \\
  Circ & 0.67 & 0.49 \\
  Moon & 0.99 & 0.98 \\
  OL & 0.37 & 0.57 \\\midrule
  Adult & 0.56 & 0.99 \\
  CH & 0.96 & 1.0 \\
  Cred & 0.67 & 1.0 \\
  GMSC & 0.38 & 1.0 \\
  MNIST & 1.0 & 1.0 \\
\end{tabular}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Generator Parameters} (Section~\ref{sec-app-grid-gen}):
  Investigates the effect of changing hyperparameters that affect the
  counterfactual outcomes during the training phase.
\item
  \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
  Investigates the effect of changing the penalty strengths in our
  proposed training objective.
\item
  \textbf{Other Parameters} (Section~\ref{sec-app-grid-train}):
  Investigates the effect of changing other training parameters,
  including the total number of generated counterfactuals in each epoch.
\end{enumerate}

@@

To measure predictive performance, we compute the accuracy and F1-score
for all models on test data (Table~\ref{tbl-acc-gen},
Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
explanatory performance, we report here our findings for the
(im)plausibility and cost of counterfactuals at test time. Since the
computation of our proposed divergence-based adaption (\(\text{IP}^*\))
is memory-intensive, we rely on the distance-based metric for the grid
searches. For the counterfactual evaluation, we draw factual samples
from the training data for the grid searches to avoid data leakage with
respect to our final results reported in the body of the paper.
Specifically, we want to avoid choosing our default hyperparameters
based on results on the test data. Since we are optimizing for
explainability, not predictive performance, we still present test
accuracy and F1-scores.

@@

We do observe strong sensitivity to certain hyperparameters, with clear
an manageable patterns. Concerning generator parameters, we firstly find
that using \emph{REVISE} to generate counterfactuals during training
typically yields the worst outcomes out of all generators, often leading
to a substantial decrease in plausibility. This finding can be
attributed to the fact that \emph{REVISE} effectively assigns the task
of learning plausible explanations from the model itself to a surrogate
VAE. In other words, counterfactuals generated by \emph{REVISE} are less
faithful to the model that \emph{ECCCo} and \emph{Generic}, and hence we
would expect them to be a less effective and, in fact, potentially
detrimental role in our training regime. Secondly, we observe that
allowing for a higher number of maximum steps \(T\) for the
counterfactual search generally yields better outcomes. This is
intuitive, because it allows more counterfactuals to reach maturity in
any given iteration. Looking in particular at the results for
\emph{Linearly Separable}, it seems that higher values for \(T\) in
combination with higher decision thresholds (\(\tau\)) yields the best
results when using \emph{ECCCo}. But depending on the degree of class
separability of the underlying data, a high decision-threshold can also
affect results adversely, as evident from the results for the
\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
here we find that CT generally fails to achieve its objective because
only a tiny proportion of counterfactuals ever reaches maturity.

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}
